Plug-and-play dual-tree algorithm runtime analysis

### 2015-01-08 bounds-files/general-runtime-bound.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

Perhaps more interesting than measures of tree imbalance is the way cover trees
are actually used in dual-tree algorithms.  Although cover trees were originally
intended for nearest neighbor search \citep*[see Algorithm
\texttt{Find-All-Nearest},][]{langford2006}, they can be adapted to a wide
variety of problems: minimum
spanning tree calculation \citep{march2010euclidean}, approximate nearest neighbor
search \citep{ram2009rank}, Gaussian processes posterior calculation
\citep{moore2014fast}, max-kernel search \citep{curtin2014dual} are some
examples.  Further, through the tree-independent dual-tree algorithm abstraction
of \citet{curtin2013tree}, other existing dual-tree algorithms can easily be
adapted for use with cover trees.

@@

    \medskip
    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
\label{alg:line:ct-dual-srmax}
    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
\label{alg:line:ct-dual-ref-recursion-start}
      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
        \STATE \texttt{BaseCase($p_q$, $p_r$)}
      \ENDFOR \label{alg:line:ct-dual-base-case-end}
      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
\label{alg:line:ct-dual-ref-set}
      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-ref-score}
      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
\label{alg:line:ct-dual-ref-recursion-end}
    \ELSE \label{alg:line:ct-dual-query-recursion-start}
      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
        \STATE $R' \gets \{ \mathscr{N}_r \in R :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-query-pruning}
        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
\label{alg:line:ct-dual-query-recursion}
      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
    \ENDIF
  \end{algorithmic}
  \caption{The standard pruning dual-tree traversal for cover trees.}
  \label{alg:cover-tree-dual}
\end{algorithm}

When using cover trees, the typical pruning dual-tree traversal is an adapted
form of the original nearest neighbor search algorithm \citep[see
\texttt{Find-All-Nearest},][]{langford2006}; this traversal is implemented in
both the cover tree reference implementation and in the more flexible {\bf
mlpack} library \citep{mlpack2013}.  The algorithm is given in Algorithm
\ref{alg:cover-tree-dual} and was originally presented by
\citet{curtin2014dual}.  Initially, it is called with the root of the query tree
and a reference set containing only the root of the reference tree.

This dual-tree recursion is a depth-first recursion in the query tree and a
breadth-first recursion in the reference tree; to this end, the recursion
maintains one query node $\mathscr{N}_q$ and a reference set $R$.  The set $R$
may contain reference nodes with many different scales; the maximum scale in the
reference set is $s_r^{\max}$ (line \ref{alg:line:ct-dual-srmax}).  Each single
recursion will descend either the query tree or the reference tree, not both;
the conditional in line \ref{alg:line:ct-dual-ref-recursion-start}, which
determines whether the query or reference tree will be recursed, is aimed at
keeping the relative scales of query nodes and reference nodes equal.

A query recursion (lines
\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end})
is straightforward: for each child $\mathscr{N}_{qc}$ of $\mathscr{N}_q$, the
node combinations $(\mathscr{N}_{qc}, \mathscr{N}_r)$ are scored for each
$\mathscr{N}_r$ in the reference set $R$.  If possible, these combinations are
pruned to form the set $R'$ (line \ref{alg:line:ct-dual-query-recursion}), and
then the algorithm recurses with $\mathscr{N}_{qc}$ and $R'$.

A reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
is similar to a query recursion, but the pruning strategy is significantly more
complicated.  We score each node combination $(\mathscr{N}_q, \mathscr{N}_r)$
for all $\mathscr{N}_r$ in $R_{r - 1}$, which is the set of children of all
nodes in $R$ that have scale $s_r^{\max}$ (that set is $R_r$).  The set of
reference nodes in $R_{r - 1}$ that were not pruned create the set $R'_{r - 1}$.
Then, this set is combined with $R \setminus R_r$---that is, each of the nodes
in $R$ that was {\em not} recursed into---to produce $R'$, and the algorithm
recurses with $\mathscr{N}_q$ and the reference set $R'$.

@@

\begin{proof}
This proof is by induction.  If $|R| = 1$, such as during the first reference
recursion, the result obviously holds.  Now consider any reference set $R$ and
assume the statement of the lemma holds for this set $R$, and define
$s_r^{\max}$ as the maximum scale of any node in $R$.  Construct the set
$R_{r - 1}$ as in line \ref{alg:line:ct-dual-ref-children} of Algorithm
\ref{alg:cover-tree-dual}.

Now, take any $\mathscr{N}_i, \mathscr{N}_j$ in $R_{r - 1}$, with points $p_i$
and $p_j$, respectively, and scales $s_i$ and $s_j$, respectively.  Clearly, if
$s_i = s_j = s_r^{\max} - 1$, then by the separation invariant $d(p_i, p_j) >
2^{s_r^{\max} - 1}$.

@@

We may therefore conclude that each point of each node in $R_{r - 1}$ is
separated by $2^{s_r^{\max} - 1}$.  Note that $R'_{r - 1} \subseteq R_{r - 1}$
and that $R \setminus R_{r - 1} \subseteq R$, and that each point of each node
is separated by $2^{s_r^{\max} - 1}$, to see that this condition holds for all
nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.

@@

These constants are directly related to the aspect ratio of the datasets;
indeed, $\eta_q / \delta_q$ is exactly the aspect ratio of $S_q$.  Further,
let us define the top and bottom levels of each tree:

\begin{itemize}
  \item The {\it top scale} of the query tree $\mathscr{T}_q$ is defined as
$s_q^T = \lceil \log_2(\eta_q) \rceil$.
  \item The {\it minimum scale} of the query tree $\mathscr{T}_r$ is defined as
$s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
  \item The top scale of the reference tree $\mathscr{T}_r$ is defined as $s_r^T
= \lceil \log_2(\eta_r) \rceil$.
  \item The minimum scale of the reference tree $\mathscr{T}_r$ is defined as
$s_r^{\min} = \lceil \log_2(\delta_r) \rceil$.
\end{itemize}

@@

\begin{equation}
\min\left(O(N), \lceil \log_2(\eta_r / \eta_q) \rceil\right).
\end{equation}
\end{lemma}

\begin{proof}
The first query recursion happens once $s_q > s_r^{\max}$.  The number of
reference recursions before the first query recursion is then bounded as the
number of levels in the reference tree between $s_r^T$ and $s_q^T$ that have at
least one explicit node.  Because there are $O(N)$ nodes in the reference tree,
the number of levels cannot be greater than $O(N)$ and thus the result holds.

The second bound holds by applying the definitions of $s_r^T$ and $s_q^T$ to the
expression $s_r^T - s_q^T$.
\end{proof}

@@

\begin{equation}
\theta = \max\left\{\min\left[O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
O(N^2)\right], \; 0\right\}.
\end{equation}
\end{lemma}

\begin{proof}
Our goal here is to count the number of reference recursions after the final
query recursion at level $s_q^{\min}$.  Because query nodes are not pruned in
this traversal, each reference recursion we are counting will be duplicated over
the whole set of $O(N)$ query nodes.  The first part of the bound follows by
observing that $s_q^{\min} - s_r^{\min} = \lceil \log_2(\delta_q) \rceil -
\lceil \log_2(\delta_r) \rceil \le \lceil \log_2(\delta_q / \delta_r) \rceil$.
The second part follows by simply observing that there are $O(N)$ reference
nodes.
\end{proof}

These two previous lemmas allow us a better understanding of what happens as the
reference set and query set become different.  Using them, we will prove general
runtime bounds for the cover tree traversal.

@@

Consider a reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end}).
The work done in the base case loop from lines
\ref{alg:line:ct-dual-base-case-start}--\ref{alg:line:ct-dual-base-case-end} is
$O(\chi | R |)$.  Define $| R^* |$ to be the largest set $|R|$ for any scale
$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
algorithm; then, it is true that $| R | \le | R^* |$.
%
Then, lines \ref{alg:line:ct-dual-ref-children} and
\ref{alg:line:ct-dual-ref-score} take $O(c_r^4 \psi | R |) \le O(c_r^4 \psi |
R^* |)$ time, because each reference node has $c_r^4$ children.  So, one
full reference recursion takes $O(c_r^4 \psi \chi | R^* |)$ time.

@@

Next, consider case \textit{(2)}.  In this situation, each query recursion
implies at least one reference recursion before another query recursion.  For
some query node $\mathscr{N}_q$, the exact number of reference recursions before
the children of $\mathscr{N}_q$ are recursed into is bounded above by
$i_t(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
exactly one level below its parent, and thus there is only one reference
recursion.  On the other hand, if $\mathscr{N}_q$ is many levels below its
parent, then it is possible that a reference recursion may occur for each level
in between; this is a maximum of $i_t(\mathscr{N}_q) + 1$.

@@

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_t(\mathscr{N}_q) + 1 = b(\mathscr{T}_q) +
O(N)
\end{equation}

@@

\begin{equation}
O(c_r^4 \psi \chi | R^* | (N + i_t(\mathscr{T}_q) + \theta)) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
\chi |R^*| (N + i_t(\mathscr{T}_q + \theta)).
\end{equation}
\end{proof}

@@

\begin{cor}
\label{cor:ct-runtime-mono}
Given the situation of Theorem \ref{thm:ct-runtime} but with $S_q = S_r = S$ so
that $c_q = c_r = c$ and $\mathscr{T}_q = \mathscr{T}_r = \mathscr{T}$, a
standard cover tree based dual-tree algorithm (Algorithm
\ref{alg:cover-tree-dual}) takes

\begin{equation}
O\left(c^4 |R^*| \chi \psi (N + i_t(\mathscr{T}))\right)
\end{equation}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The bad
performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
also have a high cover tree imbalance $i_t(\mathscr{T})$; the linear dependence of
runtime on imbalance is thus sensible.

@@

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has one unfortunate corner case: when $\delta_q > \eta_r$,
then the query tree must be entirely descended before any reference recursion.
This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
$O(N^2)$.  This is because the reference tree must be descended individually for
each query point.

The quantity $|R^*|$ bounds the amount of work that needs to be done for each
recursion. In the worst case, $|R^*|$ can be $N$. However,
dual-tree algorithms rely on branch-and-bound techniques to prune away
work (Lines \ref{alg:line:ct-dual-ref-score} and
\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
small value of $|R^*|$ will imply that the algorithm is extremely successful in
pruning away work at Line \ref{alg:line:ct-dual-ref-score} in Algorithm
\ref{alg:cover-tree-dual}. An (upper) bound on $|R^*|$ (and the algorithm's
success in pruning work) will depend on the problem and the data.  As we will
show, bounding $|R^*|$ is often possible. % In addition, we will also show that
For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
sufficient statistics \citep{moore2000anchors} can enable $O(1)$ runtime
implementations of \texttt{BaseCase()} and \texttt{Score()}.

These results hold for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
would be at least $O(N)$ using our bound, which matches the intuition that
answering $O(N)$ queries would take at least $O(N)$ time. For a particular
problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by constants
independent of $N$ and $\theta$ is no more than linear in $N$ (for large enough
$N$), then the dual-tree algorithm for that problem has a runtime linear in $N$.
Our theoretical result separates out the problem-dependent and the
problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds to get runtime bounds for any dual-tree
algorithm without requiring an analysis from scratch.

Our results are similar to that of \citet{ram2009}, but those results depend on
a quantity called the {\it constant of bichromaticity}, denoted $\kappa$, which
has unclear relation to cover tree imbalance.  The dependence on $\kappa$ is
given as $c_q^{4 \kappa}$, which is not a good bound, especially because
$\kappa$ may easily be very large.

>>>>

Perhaps more interesting than measures of tree imbalance is the way cover trees
are actually used in dual-tree algorithms.  Although cover trees were originally
intended for nearest neighbor search \citep[See Algorithm
\texttt{Find-All-Nearest},][]{langford2006}, they can be adapted to a wide
variety of problems: minimum
spanning tree calculation \citep{march2010euclidean}, approximate nearest neighbor
search \citep{ram2009rank}, Gaussian processes posterior calculation
\citep{moore2014fast}, and max-kernel search \citep{curtin2014dual} are some
examples.  Further, through the tree-independent dual-tree algorithm abstraction
of \citet{curtin2013tree}, other existing dual-tree algorithms can easily be
adapted for use with cover trees.

@@

    \medskip
    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
\label{alg:line:ct-dual-srmax}
    \IF{$(s_q < s^{\max}_r)$}
\label{alg:line:ct-dual-ref-recursion-start}
      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
        \STATE \texttt{BaseCase($p_q$, $p_r$)}
      \ENDFOR \label{alg:line:ct-dual-base-case-end}
      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
\label{alg:line:ct-dual-ref-set}
      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-ref-score}
      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
\label{alg:line:ct-dual-ref-recursion-end}
    \ELSE \label{alg:line:ct-dual-query-recursion-start}
      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
        \STATE $R' \gets \{ \mathscr{N}_r \in R :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-query-pruning}
        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
\label{alg:line:ct-dual-query-recursion}
      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
    \ENDIF
  \end{algorithmic}
  \caption{The standard pruning dual-tree traversal for cover trees.}
  \label{alg:cover-tree-dual}
\end{algorithm}

When using cover trees, the typical pruning dual-tree traversal is an adapted
form of the original nearest neighbor search algorithm \citep[see
\texttt{Find-All-Nearest},][]{langford2006}; this traversal is implemented in
both the cover tree reference implementation and in the more flexible {\bf
mlpack} library \citep{mlpack2013}.  The problem-independent traversal is given
in Algorithm \ref{alg:cover-tree-dual} and was originally presented by
\citet{curtin2014dual}.  Initially, it is called with the root of the query tree
and a reference set $R$ containing only the root of the reference tree.

This dual-tree recursion is a depth-first recursion in the query tree and a
breadth-first recursion in the reference tree; to this end, the recursion
maintains one query node $\mathscr{N}_q$ and a reference set $R$.  The set $R$
may contain reference nodes with many different scales; the maximum scale in the
reference set is $s_r^{\max}$ (line \ref{alg:line:ct-dual-srmax}).  Each single
recursion will descend either the query tree or the reference tree, not both;
the conditional in line \ref{alg:line:ct-dual-ref-recursion-start}, which
determines whether the query or reference tree will be recursed, is aimed at
keeping the relative scales of query nodes and reference nodes close.

A query recursion (lines
\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end})
is straightforward: for each child $\mathscr{N}_{qc}$ of $\mathscr{N}_q$, the
node combinations $(\mathscr{N}_{qc}, \mathscr{N}_r)$ are scored for each
$\mathscr{N}_r$ in the reference set $R$.  If possible, these combinations are
pruned to form the set $R'$ (line \ref{alg:line:ct-dual-query-recursion}) by
checking the output of the \texttt{Score()} function, and then the algorithm
recurses with $\mathscr{N}_{qc}$ and $R'$.

A reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
is similar to a query recursion, but the pruning strategy is significantly more
complicated.  Given $R$, we calculate $R_r$, which is the set of nodes in $R$
that have scale $s_r^{\max}$.  Then, for each node $\mathscr{N}_r$ in the set of
children of nodes in $R_r$, the node combinations $(\mathscr{N}_q,
\mathscr{N}_r)$ are scored and pruned if possible.  Those reference nodes that
were not pruned form the set $R'_{r - 1}$.  Then, this set is combined with $R
\setminus R_r$---that is, each of the nodes in $R$ that was {\em not} recursed
into---to produce $R'$, and the algorithm recurses with $\mathscr{N}_q$ and the
reference set $R'$.

@@

\begin{proof}
This proof is by induction.  If $|R| = 1$, such as during the first reference
recursion, the result obviously holds.  Now consider any reference set $R$ and
assume the statement of the lemma holds for this set $R$, and define
$s_r^{\max}$ as the maximum scale of any node in $R$.  Construct the set
$R_{r - 1}$ as in line \ref{alg:line:ct-dual-ref-children} of Algorithm
\ref{alg:cover-tree-dual}; if $| R_{r - 1} | \le 1$, then $R_{r - 1}$ satisfies
the desired property.

Otherwise, take any $\mathscr{N}_i, \mathscr{N}_j$ in $R_{r - 1}$, with points
$p_i$ and $p_j$, respectively, and scales $s_i$ and $s_j$, respectively.
Clearly, if $s_i = s_j = s_r^{\max} - 1$, then by the separation invariant
$d(p_i, p_j) > 2^{s_r^{\max} - 1}$.

@@

We may therefore conclude that each point of each node in $R_{r - 1}$ is
separated by $2^{s_r^{\max} - 1}$.  Note that $R'_{r - 1} \subseteq R_{r - 1}$
and that $R \setminus R_{r - 1} \subseteq R$ in order to see that this condition
holds for all nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.

@@

These constants are directly related to the aspect ratio of the datasets;
indeed, $\eta_q / \delta_q$ is exactly the aspect ratio of $S_q$.  Further,
let us define and bound the top and bottom levels of each tree:

\begin{itemize}
  \item The {\it top scale} $s_q^T$ of the query tree $\mathscr{T}_q$ is such
that as $\lceil \log_2(\eta_q) \rceil - 1 \le s_q^T \le \lceil \log_2(\eta_q)
\rceil$.
  \item The {\it minimum scale} of the query tree $\mathscr{T}_r$ is defined as
$s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
  \item The top scale $s_r^T$ of the reference tree $\mathscr{T}_r$ is such that
as $\lceil \log_2(\eta_r) \rceil - 1 \le s_r^T \le \lceil \log_2(\eta_r)
\rceil$.
  \item The minimum scale of the reference tree $\mathscr{T}_r$ is defined as
$s_r^{\min} = \lceil \log_2(\delta_r) \rceil$.
\end{itemize}

@@

\begin{equation}
\min\left(O(N), \log_2(\eta_r / \eta_q) - 1\right).
\end{equation}
\end{lemma}

\begin{proof}
The first query recursion happens once $s_q \ge s_r^{\max}$.  The number of
reference recursions before the first query recursion is then bounded as the
number of levels in the reference tree between $s_r^T$ and $s_q^T$ that have at
least one explicit node.  Because there are $O(N)$ nodes in the reference tree,
the number of levels cannot be greater than $O(N)$ and thus the result holds.

The second bound holds by applying the definitions of $s_r^T$ and $s_q^T$ to the
expression $s_r^T - s_q^T - 1$:

\begin{eqnarray}
s_r^T - s_q^T - 1 &\le& \lceil \log_2(\eta_r) \rceil - (\lceil \log_2(\eta_q)
\rceil - 1) - 1 \\
&\le& \log_2(\eta_r) + 1 - \log_2(\eta_q)
\end{eqnarray}

\noindent which gives the statement of the lemma after applying logarithmic
identities.
\end{proof}

@@

\begin{equation}
\theta = \max\left\{\min\left[O(N \log_2(\delta_q / \delta_r)),
O(N^2)\right], \; 0\right\}.
\end{equation}
\end{lemma}

\begin{proof}
Our goal here is to count the number of reference recursions after the final
query recursion at level $s_q^{\min}$; the first of these reference recursions
is at scale $s_r^{\max} = s_q^{\min}$.  Because query nodes are not pruned in
this traversal, each reference recursion we are counting will be duplicated over
the whole set of $O(N)$ query nodes.  The first part of the bound follows by
observing that
$s_q^{\min} - s_r^{\min} \le \lceil \log_2(\delta_q) \rceil -
\lceil \log_2(\delta_r) \rceil - 1 \le \log_2(\delta_q / \delta_r)$.

The second part follows by simply observing that there are $O(N)$ reference
nodes.
\end{proof}

These two previous lemmas allow us a better understanding of what happens as the
reference set and query set become different.  Lemma \ref{lem:extcase1} shows
that the number of extra recursions caused by a reference set with larger
pairwise distances than the query set ($\eta_r$ larger than $\eta_q$) is modest;
on the other hand, Lemma \ref{lem:extcase3} shows that for each extra level in
the reference tree below $s_q^{\min}$, $O(N)$ extra recursions are required.
Using these lemmas and this intuition, we will prove general runtime bounds for
the cover tree traversal.

@@

Consider a reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end}).
Define $R^*$ to be the largest set $R$ for any scale
$s_r^{\max}$ and any query node $\mathscr{N}_q$ during the course of the
algorithm; then, it is true that $| R | \le | R^* |$.  The work done in the base
case loop from lines
\ref{alg:line:ct-dual-base-case-start}--\ref{alg:line:ct-dual-base-case-end} is
thus $O(\chi | R |) \le O(\chi | R^*|)$.
%
Then, lines \ref{alg:line:ct-dual-ref-children} and
\ref{alg:line:ct-dual-ref-score} take $O(c_r^4 \psi | R |) \le O(c_r^4 \psi |
R^* |)$ time, because each reference node has up to $c_r^4$ children.  So, one
full reference recursion takes $O(c_r^4 \psi \chi | R^* |)$ time.

@@

Next, consider case \textit{(2)}.  In this situation, each query recursion
implies at least one reference recursion before another query recursion.  For
some query node $\mathscr{N}_q$, the exact number of reference recursions before
the children of $\mathscr{N}_q$ are recursed into is bounded above by
$i_n(\mathscr{N}_q) + 1$: if $\mathscr{N}_q$ has imbalance $0$, then it is
exactly one level below its parent, and thus there is only one reference
recursion.  On the other hand, if $\mathscr{N}_q$ is many levels below its
parent, then it is possible that a reference recursion may occur for each level
in between; this is a maximum of $i_n(\mathscr{N}_q) + 1$.

@@

\begin{equation}
\sum_{\mathscr{N}_q \in \mathscr{T}_q} i_n(\mathscr{N}_q) + 1 = i_t(\mathscr{T}_q) +
O(N)
\end{equation}

@@

\begin{equation}
O\left(c_r^4 |R^*| \psi \chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right)
+ O\left(c_r^4 |R^*| \psi N\right) \sim O\left(c_r^4 |R^*| \psi
\chi \left(N + i_t(\mathscr{T}_q) + \theta\right)\right).
\end{equation}
\end{proof}

@@

\begin{cor}
\label{cor:ct-runtime-mono}
Given the situation of Theorem \ref{thm:ct-runtime} but with $S_q = S_r = S$ so
that $c_q = c_r = c$ and $\mathscr{T}_q = \mathscr{T}_r = \mathscr{T}$, a
dual-tree algorithm using the standard cover tree traversal (Algorithm
\ref{alg:cover-tree-dual}) takes

\begin{equation}
O\left(c^4 |R^*| \chi \psi \left(N + i_t(\mathscr{T})\right)\right)
\end{equation}

@@

An intuitive understanding of these bounds is best achieved by first considering
the monochromatic case (this case arises, for instance, in all-nearest-neighbor
search).  The linear dependence on $N$ arises from the fact that all query nodes
must be visited.  The dependence on the reference tree, however, is encapsulated
by the term $c^4 |R^*|$, with $|R^*|$ being the maximum size of the reference
set $R$; this value must be derived for each specific problem.  The bad
performance of poorly-behaved datasets with large $c$ (or, in the worst case, $c
\sim N$) is then captured in both of those terms.  Poorly-behaved datasets may
also have a high cover tree imbalance $i_t(\mathscr{T})$; the linear dependence of
runtime on imbalance is thus sensible for well-behaved datasets.

@@

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has one unfortunate corner case: when $\delta_q > \eta_r$,
then the query tree must be entirely descended before any reference recursion.
This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
$O(N^2)$ (See Lemma \ref{lem:extcase3}).  This is because the reference tree
must be descended individually for each query point.

The quantity $|R^*|$ bounds the amount of work that needs to be done for each
recursion. In the worst case, $|R^*|$ can be $N$. However,
dual-tree algorithms rely on branch-and-bound techniques to prune away
work (Lines \ref{alg:line:ct-dual-ref-score} and
\ref{alg:line:ct-dual-query-pruning} in Algorithm \ref{alg:cover-tree-dual}). A
small value of $|R^*|$ will imply that the algorithm is extremely successful in
pruning away work.  An (upper) bound on $|R^*|$ (and the algorithm's
success in pruning work) will depend on the problem and the data.  As we will
show, bounding $|R^*|$ is often possible. % In addition, we will also show that
For many dual-tree algorithms, $\chi \sim \psi \sim O(1)$; often, cached
sufficient statistics \citep{moore2000anchors} can enable $O(1)$ runtime
implementations of \texttt{BaseCase()} and \texttt{Score()}.

These results hold for any dual-tree algorithm regardless of the problem. Hence,
%for any pairwise statistical problem, the runtime of the dual-tree algorithm (of
the runtime of any dual-tree algorithm
%that problem)
would be at least $O(N)$ using our bound, which matches the intuition that
answering $O(N)$ queries would take at least $O(N)$ time. For a particular
problem and data, if $c_r$, $|R^*|$, $\chi$, $\psi$ are bounded by constants
independent of $N$ and $\theta$ is no more than linear in $N$ (for large enough
$N$), then the dual-tree algorithm for that problem has a runtime linear in $N$.
Our theoretical result separates out the problem-dependent and the
problem-independent elements of the runtime bound, which allows us to simply
plug in the problem-dependent bounds in order to get runtime bounds for any
dual-tree algorithm without requiring an analysis from scratch.

Our results are similar to that of \citet{ram2009}, but those results depend on
a quantity called the {\it constant of bichromaticity}, denoted $\kappa$, which
has unclear relation to cover tree imbalance.  The dependence on $\kappa$ is
given as $c_q^{4 \kappa}$, which is not a good bound, especially because
$\kappa$ may be much greater than $1$ in the bichromatic case (where $S_q \ne
S_r$).
